---
title: "W2: Working with data structures"
format: 
  revealjs:
    smaller: true
    scrollable: true
    echo: true
    embed-resources: true
    drop:
      engine: webr
      button: true
revealjs-plugins:
  - drop
output-location: fragment
---

## Data types

-   **Numeric**: 18, -21, 65, 1.25
-   **Character**: "ATCG", "Whatever", "948-293-0000"
-   **Logical**: TRUE, FALSE

## Data structures

**Data structures** stores information about data types.

. . .

**Vector** is a ordered collection of a data type. Each *element* of a vector contains a data type, and all elements of a vector must be the same type, such as numeric, character, or logical.

. . .

So, you can have **numeric vector**, **character vector**, and **logical vector**.

::: notes
There is no limit on how big a vector can be, as long the memory use of it is within the computer's memory (RAM).

We can now store a vast amount of information in a vector, and assign it to a single variable. We can now use operations and functions on a vector, modifying many elements within the vector at once!
:::

## Defining a vector

. . .

```{r}
staff = c("chris", "shasta", "jeff")
chrNum = c(2, 3, 1)
```

## Using operations on vectors

. . .

You can use operations and functions on vectors.

. . .

```{r}
chrNum = chrNum * 3
chrNum 
```

. . .

Multiplication has a new meaning when used on a numeric vector and numeric data type! This is called **operator overloading**.

. . .

😮

## Operator overloading

How about numeric vector multiplied by numeric vector?

```{r}
chrNum * c(2, 2, 0)
```

. . .

a numeric vector added to a character vector:

```{r}
#chrNum + staff
```

## Functions on vectors

How many elements are in this vector?

```{r}
length(staff)
```

## Subsetting vectors

We subset vectors using the bracket `[ ]` operation.

. . .

Inside the bracket is either a single numeric value or an a **numerical indexing vector** containing numerical values. They dictate which elements of the vector to return.

. . .

```{r}
staff[2]
```

. . .

```{r}
staff[c(1, 2)]
```

. . .

```{r}
small_staff = staff[c(1, 2)]
small_staff
```

## Subsetting vectors

Alternatively, instead of using numerical indexing vectors, we can use a **logical indexing vector**.

. . .

The logical indexing vector must be the *same length* as the vector to be subsetted, with `TRUE` indicating an element to keep, and `FALSE` indicating an element to drop.

. . .

```{r}
staff[c(TRUE, FALSE, FALSE)]
```

. . .

```{r}
staff[c(TRUE, TRUE, FALSE)]
```

. . .

```{r}
small_staff = staff[c(TRUE, TRUE, FALSE)]
small_staff
```

## Explicit vs. implicit subsetting

. . .

-   **Explicit subsetting**: Given a length 10 vector of people's ages, and subset to the 1st, 3rd, and 9th elements.

. . .

```{r}
age = c(89, 70, 64, 90, 66, 71, 55, 60, 30, 16)
```

. . .

```{r}
age[c(1, 5, 9)]
```

. . .

```{r}
age[c(TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE)]
```

. . .

-   **Implicit subsetting**: Given a length 10 vector of people's ages, and subset to elements greater than age 50.

## Implicit subsetting

We don't know which elements to subset off the top of our head! If we know which elements are \> 50, then we can give the elements for an explicit subset.

. . .

Use a **comparison operator** to create a logical indexing vector:

. . .

```{r}
age > 50
```

. . .

Can you guess what we do next?

. . .

```{r}
age[age > 50]
```

## Implicit subsetting

Subset a vector **implicitly**, in 3 steps:

. . .

1.  Come up with a criteria for subsetting: "I want to subset to values greater than 50".

. . .

2.  Use a **comparison operator** to create a **logical indexing vector** that fits this criteria.

```{r}
age > 50
```

. . .

3.  Use this logical indexing vector to subset.

```{r}
age[age > 50]
```

. . .

Alternatively,

```{r}
idx = age > 50
age[idx]
```

## Comparison operators:

`<` less than

`<=` less or equal than

`>` greater than

`>=` greater than or equal to

`==` equal to

`!=` not equal to

. . .

Subset `staff` to not have "chris" in it.

. . .

```{r}
staff != "chris"
```

. . .

```{r}
staff[staff != "chris"]
```

## Dataframes

. . .

A **dataframe** is a spreadsheet such that each column must have the same data type; a bunch of vectors organized as columns.

. . .

```{r}
load(url("https://github.com/fhdsl/Intro_to_R/raw/main/classroom_data/CCLE.RData"))
nrow(metadata)
```

. . .

```{r}
ncol(metadata)
```

. . .

```{r}
dim(metadata)
```

. . .

`head(metadata)`

`View(metadata)`

## Column names of dataframes

The **column name** is a character vector that corresponds to the columns of the dataframe.

. . .

```{r}
names(metadata)
```

. . .

## Subsetting a column from a dataframe

. . .

The `dataframe$column_name` operation selects for a column by its column name and returns the column as a vector:

. . .

```{r}
metadata$OncotreeLineage[1:5]
```

. . .

```{r}
metadata$Age[1:5]
```

. . .

Challenge!

```{r}
metadata$Age[metadata$OncotreeLineage == "Myeloid"]
```

## Subsetting columns and rows from a dataframe

`dataframe[row_idx, col_idx]` subsets the dataframe by a row indexing vector `row_idx`, and a column indexing vector `col_idx`.

. . .

```{r}
metadata[1:5, c(1, 3)]
```

. . .

We can refer to the column names directly:

```{r}
metadata[1:5, c("ModelID", "CellLineName")]
```

## Subsetting columns and rows from a dataframe

We can leave the column index or row index empty to just subset columns or rows.

. . .

```{r}
metadata[1:5, ]
```

. . .

```{r}
metadata[, c("ModelID", "CellLineName")]
```
